---
title: "Light Calcification"
author: "Jill Ashey"
date: "2026-05-08"
output: html_document
---

This script analyzed light calcification data from the heat stress experiment. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(emmeans)
library(lme4)
library(lmerTest)
library(lsmeans)
library(car)
```

Run this before running any ANOVAs
```{r}
options(contrasts = c("contr.sum", "contr.poly"))
```

Read in data 
```{r}
calc <- read.csv("../data/Physiology/Calcification_data_experiment.csv", header = T)
```

Calculate incubation time and delta alkalinity based on the initial water sample
```{r}
calc <- calc %>%
  mutate(
    start = hm(Time_start),
    end   = hm(Time_end),
    elapsed_mins  = as.numeric(end - start, units = "mins"),
    elapsed_hours = elapsed_mins / 60) %>%
  mutate(Delta_TA_raw = Starting_alk_value - Alk_avg)
```

Calculate blanks 
```{r}
blank_delta_ta <- calc %>%
  filter(SampleID == "blank" | Notes == "blank") %>%
  group_by(Timepoint) %>%
  summarise(mean_blank_delta_ta = mean(Delta_TA_raw, na.rm = TRUE))# %>%
  #pull(mean_blank_delta_ta)
```

Subtract blank and calculate net calcification
```{r}
# Correct Delta_TA by subtracting blank
calc <- calc %>%
  left_join(blank_delta_ta, by = "Timepoint") %>% 
  mutate(
    Delta_TA_corr = Delta_TA_raw - mean_blank_delta_ta,
    # Assuming seawater density 1.028 g/mL and volume in mL to L, and no coral weight (thus calc mg/L/hr)
    Net_Calcification = (Delta_TA_corr / 2) * (Sample_Volume_mL / 1000) * 1.028 / elapsed_hours) %>%
  filter(!is.na(Net_Calcification) & !is.nan(Net_Calcification))
```

Read in surface area data and join with calc df 
XXXXXXX

Normalize by SA--setting SA to 10 for now until we get actual surface area data from fraggrameter or wax dipping
```{r}
calc <- calc %>%
  mutate(Net_Calcification_cm2 = Net_Calcification / 10)
```

Read in sample metadata and join with calc df by sample ID
```{r}
meta <- read.csv("../data/sample_manifest.csv", header = T)

calc_meta <- calc %>%
  left_join(meta, by = "SampleID")
```

Calculate mean, sem, and sd 
```{r}
calc_summary <- calc_meta %>%
  group_by(Timepoint, Phenotype, Species, Treatment) %>%
  summarise(
    mean_calc = mean(Net_Calcification_cm2, na.rm = TRUE),
    sd_calc = sd(Net_Calcification_cm2, na.rm = T),
    se_calc = sd(Net_Calcification_cm2, na.rm = T) / sqrt(length(Net_Calcification_cm2)),
    .groups = "drop") %>%
  drop_na()
```

Plot
```{r}
ggplot(calc_summary, aes(x = Timepoint, y = mean_calc, color = Phenotype, group = Phenotype)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey", size = 1) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_calc - se_calc, ymax = mean_calc + se_calc, color = Phenotype),
                width = 0.3, linewidth = 0.8) +  # color matches points/lines
  facet_wrap(~Species+Treatment, scales = "fixed") +
  scale_color_manual(
    values = c("Resistant" = "#4b294fff", "Susceptible" = "#f3a17dff")
  ) +
  theme_classic(base_size = 16) +  # bigger base size
  labs(
    x = "",
    y = "Net calcification (umol/hr/cm2)",
    color = "Phenotype"
  ) +
theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    strip.background = element_rect(fill = "white", color = "black", linewidth = 1),
    strip.text = element_text(face = "italic", size = 16),
    #legend.position = c(0.85, 0.83),  # adjust x, y for placement
    legend.position = "right",
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.key = element_blank(),    
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 13),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 13, color = "black"),
    axis.text.y = element_text(size = 13, color = "black"),
    axis.title.y = element_text(size = 16)
  )
```

```{r}
library(ggplot2)
library(dplyr)

# 1. Calculate means and standard errors
summary_data <- calc_meta %>%
  filter(!is.na(Species) & !is.na(Phenotype) & !is.na(Treatment)) %>%
  group_by(Timepoint, Species, Phenotype, Treatment) %>%
  summarise(
    mean_calc = mean(Net_Calcification_cm2, na.rm = TRUE),
    se_calc = sd(Net_Calcification_cm2, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

# 2. Plot the trends
ggplot(summary_data, aes(x = Timepoint, y = mean_calc, color = Treatment, group = Treatment)) +
  geom_line(linewidth = 1, position = position_dodge(0.2)) +
  geom_errorbar(aes(ymin = mean_calc - se_calc, ymax = mean_calc + se_calc), 
                width = 0.1, position = position_dodge(0.2)) +
  geom_point(size = 3, position = position_dodge(0.2)) +
  facet_grid(Species ~ Phenotype) +
  scale_color_manual(values = c("Ambient" = "#56B4E9", "High" = "#D55E00")) +
  theme_bw() +
  labs(
    title = "Changes in Net Calcification Across Timepoints",
    x = "Timepoint",
    #y = expression(Mean~Net~Calcification \pm SE),
    color = "Treatment"
  ) +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "bottom"
  )
```

```{r}
library(ggplot2)
library(dplyr)

# 1. Clean the data (remove blanks/NAs in key columns)
plot_data <- calc_meta %>%
  filter(!is.na(Species) & !is.na(Phenotype) & !is.na(Treatment))

# 2. Generate the plot
ggplot(plot_data, aes(x = Timepoint, y = Net_Calcification_cm2, color = Phenotype)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(0.8)) +
  geom_point(aes(group = Treatment), position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.8), size = 1.5, alpha = 0.6) +
  facet_grid(Species ~ Treatment) + 
  scale_color_manual(
    values = c("Resistant" = "#4b294fff", "Susceptible" = "#f3a17dff")) +
  #scale_fill_manual(values = c("Ambient" = "#56B4E9", "High" = "#D55E00")) + # Custom color palette
  theme_bw() +
  labs(
    title = "Net Calcification by Species, Phenotype, Treatment, and Timepoint",
    x = "Timepoint",
    y = expression(Net~Calcification~(mg~cm^{-2}~hr^{-1})) # Formats units nicely
  ) +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "bottom"
  )
```

Stats! 
```{r}
stats_data <- calc_meta %>%
  filter(!is.na(Species) & !is.na(Phenotype) & !is.na(Treatment)) %>%
  mutate(
    Timepoint = as.factor(Timepoint),
    Phenotype = as.factor(Phenotype),
    Treatment = as.factor(Treatment),
    ColonyID  = as.factor(ColonyID) # Useful if tracking individuals
  )

## Pcomp 
pc_data <- stats_data %>% filter(Species == "Porites compressa")
pc_mixed_model <- lmer(Net_Calcification_cm2 ~ Treatment * Timepoint * Phenotype + (1|ColonyID), data = pc_data)
anova(pc_mixed_model)

## Mcap 
mc_data<- stats_data %>% filter(Species == "Montipora capitata")
mc_mixed_model <- lmer(Net_Calcification_cm2 ~ Treatment * Timepoint * Phenotype + (1|ColonyID), data = mc_data)
anova(mc_mixed_model)
```

