---
title: "Discrete measurements"
author: "Jill Ashey"
date: "2026-05-08"
output: html_document
---

This script plots data from discrete measurements (temperature, pH, salinity, light). 

Set up 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

Load packages
```{r}
library(tidyverse)
library(stringr)
library(readxl)
library(purrr)
library(lubridate)
library(ggplot2)
library(broom)
library(cowplot)
library(ggh4x)
```

Read in daily measurement csv 
```{r}
daily <- read.csv("../data/environmental/Discrete_Measurements.csv")
daily$Date <- as.Date(daily$Date, format="%m/%d/%y")
```

Change data to long format 
```{r}
daily_long <- daily %>% 
  pivot_longer(cols=Temperature_C:Light_PAR,
  names_to = "metric",
  values_to = "value")
```

Filter by dates of interest (ie experimental period)
XXXXX

Plot by tank over time 
```{r}
daily_tanks <- daily_long %>%
  ggplot(aes(x=Date, y=value, colour=Tank))+
  geom_point(size=2)+
  xlab("Date")+
  facet_grid(metric ~ ., scales = "free")+
  theme_bw(); daily_tanks

ggsave(filename="../output/environmental/tank_daily_measurements.png", plot=daily_tanks, dpi=300, width=6, height=8, units="in")
```

Plot by treatment summarizing tanks within each group over time 
```{r}
daily_treatment <- daily_long %>%
  ggplot(aes(x=Date, y=value, colour=Treatment, group=interaction(Treatment, Date)))+
  geom_point()+
  #geom_boxplot()+
  scale_colour_manual(values = c("Ambient" = "blue", "High" = "red")) +
  xlab("Date")+
  facet_grid(metric ~ ., scales = "free")+
  theme_bw(); daily_treatment

ggsave(filename="../output/environmental/treatment_daily_measurements.png", plot=daily_treatment, dpi=300, width=6, height=8, units="in")
```

Plot summary of treatments
```{r}
daily_treatment_summary <- daily_long %>%
  ggplot(aes(x = Treatment, y = value, colour = Treatment)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.15, height = 0), alpha = 0.6, size = 1.8) +
  scale_colour_manual(values = c("Ambient" = "blue", "High" = "red")) +
  ylab("Value") +
  #ggtitle("A")+
  ggh4x::facet_wrap2(~ metric, scales = "free_y", nrow = 2, axes = "margins") +  # <- x-axis only on outer/bottom
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.placement = "outside",
    strip.background = element_blank()
  ); daily_treatment_summary

ggsave(filename="../output/environmental/summary_daily_measurements.png", plot=daily_treatment_summary, dpi=300, width=7, height=6, units="in")
```

Test for differences between tanks 
```{r}
# Temperature
daily %>%
  aov(Temperature_C~Tank, data=.)%>%
  summary()

# pH
daily %>%
  aov(pH_NBS~Tank, data=.)%>%
  summary()

# Salinity 
daily %>%
  aov(Salinity_psu~Tank, data=.)%>%
  summary()

# Light
daily %>%
  aov(Light_PAR~Tank, data=.)%>%
  summary()

# Flow
daily %>%
  aov(Flow_mL_per_5s~Tank, data=.)%>%
  summary()
```

Test for differences between treatments
```{r}
# Temperature
daily %>%
  aov(Temperature_C~Treatment, data=.)%>%
  summary()

# pH
daily %>%
  aov(pH_NBS~Treatment, data=.)%>%
  summary()

# Salinity 
daily %>%
  aov(Salinity_psu~Treatment, data=.)%>%
  summary()

# Light
daily %>%
  aov(Light_PAR~Treatment, data=.)%>%
  summary()

# Flow
daily %>%
  aov(Flow_mL_per_5s~Treatment, data=.)%>%
  summary()
```
