---
title: "PI curves"
author: "Jill Ashey"
date: "2026-05-08"
output: html_document
---

This script analyzes PI curve data generated during the heat stress experiment. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("devtools")
#devtools::install_github("colin-olito/LoLinR")

library(tidyverse)
library(nls.multstart)
library(broom)
library(LoLinR)
library(readr)
library(lubridate)
library(fuzzyjoin)
library(stringr)
library(purrr)
library(future)
library(furrr)
library(minpack.lm)
library(lme4)
library(lmerTest)
library(nlme)
library(emmeans)
```

## Initial 

### Prep data 

Load run metadata 
```{r}
run_meta <- read_csv("../data/Physiology/PI_curves/pi_curve_metadata_experiment_run.csv")%>%
  mutate(
    Start = as_datetime(Date) + hours(hour(Start.time)) + minutes(minute(Start.time)),
    Stop  = as_datetime(Date) + hours(hour(Stop.time)) + minutes(minute(Stop.time)))
```

Load sample metadata
```{r}
sample_meta <- read_csv("../data/Physiology/PI_curves/pi_curve_metadata_experiment_sample.csv")

# Make sure Chamber well # is numeric
sample_meta <- sample_meta %>%
  mutate(`Chamber` = as.numeric(`Chamber`))
```

List all raw files
```{r}
files <- list.files(path = "../data/Physiology/PI_curves/raw_data/20260502", pattern = "*.csv", full.names = TRUE)
```

Join O2 data with metadata files based on start and stop times 
```{r}
all_data <- map_dfr(files, function(f) {
  # Extract channel number from filename, e.g. Ch1 → 1
  channel <- str_extract(f, "Ch\\d+") %>% str_remove("Ch") %>% as.numeric()
  
  # Read the file
  raw <- read_csv(f) %>%
    mutate(
      # Parse datetime (adjust format if needed)
      DateTime = dmy_hms(Date),
      Channel = channel
    )
  
  # Join to run metadata by fuzzy time match
  raw_joined <- fuzzy_left_join(
    raw,
    run_meta,
    by = c("DateTime" = "Start", "DateTime" = "Stop"),
    match_fun = list(`>=`, `<=`)
  )
  
  # Join to sample metadata by Date, Run #, and Channel
  raw_joined <- raw_joined %>%
    left_join(
      sample_meta,
      by = c(
        "Date.y" = "Date",
        "Run" = "Run",
        "Channel" = "Chamber"
      )
    )
  
  return(raw_joined)
})
```

Look at joined data and save
```{r}
all_data %>% count(is.na(SampleID))
all_data %>% count(is.na(Run))

write_csv(all_data, "../output/Physiology/PI_curves/all_joined_data_May2026_phys.csv")
```

### Run localized linear regressions 

The interval fitting code will take about an hour to run and is computationally intense. If needed, run as a background job or on a remote server.

Make list of all SampleID / Run combinations (including blanks)
```{r}
colony_runs <- all_data %>%
  filter(!is.na(SampleID)) %>%
  distinct(SampleID, Run)
```

Calculate interval slopes for the blanks
```{r}
blank_data <- all_data %>%
  filter(SampleID == "blank")

blank_intervals <- blank_data %>%
  group_by(Run, Light_Value) %>%
  filter(!is.na(SampleID)) %>%
  group_split()
```

Fit regressions and extract slopes for each blank interval
```{r}
fit_reg <- function(df) {
  rankLocReg(
    xall = as.numeric(df$`Delta T [min]`),
    yall = df$Oxygen,
    alpha = 0.2,
    method = "pc",
    verbose = FALSE
  )
}

blank_fits <- map(blank_intervals, fit_reg)
blank_slopes <- map_dbl(blank_fits, ~ pluck(., "allRegs", "b1", 1))
blank_summary <- map2_dfr(
  blank_intervals,
  blank_slopes,
  ~ tibble(
    Light_Value = unique(.x$Light_Value),
    Blank_Slope = .y
  )
)
```

Average blank slopes per light value, across all blanks/runs
```{r}
avg_blank_slopes <- blank_summary %>%
  group_by(Light_Value) %>%
  summarise(Avg_Blank_Slope = mean(Blank_Slope, na.rm = TRUE), .groups = "drop")
```

Run interval fittings with blank correction
```{r}
# Function: blank correction + slope extraction + save/return interval summary
fit_colony_intervals_avgblank <- function(sample_id, run_num) {
  # Subset sample data
  sample_data <- all_data %>%
    filter(SampleID == sample_id, Run == run_num)
  if (nrow(sample_data) == 0) return(NULL)
  
  # Split data by light step
  intervals <- sample_data %>%
    group_by(Light_Value) %>%
    group_split()
  
  # Fit regressions for each interval
  fits <- map(intervals, fit_reg)
  micromol.L.s <- map_dbl(fits, ~ pluck(., "allRegs", "b1", 1))
  
  # Detailed interval summary
  interval_df <- map2_dfr(
    intervals,
    seq_along(intervals),
    ~ tibble(
      SampleID   = sample_id,
      Run         = run_num,
      Interval    = .y,
      Light_Value = unique(.x$Light_Value),
      N_points    = nrow(.x),
      O2_start    = head(.x$Oxygen, 1),
      O2_end      = tail(.x$Oxygen, 1),
      DeltaT_min  = min(.x$`Delta T [min]`),
      DeltaT_max  = max(.x$`Delta T [min]`),
      O2_min      = min(.x$Oxygen, na.rm = TRUE),
      O2_max      = max(.x$Oxygen, na.rm = TRUE),
      Slope       = micromol.L.s[.y]
    )
  ) %>%
    left_join(avg_blank_slopes, by = "Light_Value") %>%
    mutate(
      Blank_Corrected_Slope = Slope - Avg_Blank_Slope
    )
  
  # Write file (outside of the pipeline!)
  fn <- paste0("../output/Physiology/PI_curves/intervals/intervals_", sample_id, "_run", run_num, "_May2026_experiment_TP0.csv")
  write_csv(interval_df, fn)
  
  interval_df
}

# Run for all samples and save big CSV
plan(multisession)

all_interval_summaries <- future_pmap_dfr(
  list(colony_runs$SampleID, colony_runs$Run),
  fit_colony_intervals_avgblank
)

write_csv(all_interval_summaries, "../output/Physiology/PI_curves/intervals/all_interval_summaries_May2026_experiment_TP0.csv")
```



^^ come back to this 

Read in TP1 and TP2 data 
```{r}
pi_tp1 <- read.csv("../output/Physiology/PI_curves/all_interval_summaries_May2026_experiment_TP1.csv")
```

Read in sample metadata and join with calc df by sample ID
```{r}
meta <- read.csv("../data/sample_manifest.csv", header = T)

pi_tp1 <- pi_tp1 %>%
  left_join(meta, by = "SampleID") %>%
  left_join(sample_meta, by = "SampleID")
```

normalize 
```{r}
pi_tp1_norm <- pi_tp1 %>%
  mutate(
    Chamber_Volume_L = ((Chamber_Volume_mL -100) / 1000),  # mL → L
    Slope_norm = (Blank_Corrected_Slope * Chamber_Volume_L / 10) * 60) %>% # units are umol O2 cm-2 h-1
  filter(!is.na(Phenotype))
```

Plot across light values 
```{r}
# Plot PI curves across light values 
ggplot(pi_tp1_norm, aes(x = Light_Value, y = Slope_norm, color = Phenotype)) +
  geom_hline(yintercept = 0, alpha = 0.5, linetype = "dashed") +
  geom_point() +
  geom_line(aes(group = SampleID)) + 
  scale_color_manual(
    values = c("Resistant" = "#4b294fff", "Susceptible" = "#f3a17dff")) +
  facet_wrap(~Species+Treatment, scales = "fixed") +
  theme_classic(base_size = 14) +
  labs(x = "Light (µmol photons m-2 s-1)", y = "umol O2 cm-2 h-1", color = "Phenotype")

```

Remove 12-A1
```{r}
pi_tp1_norm <- pi_tp1_norm %>%
  filter(!SampleID == "MC-12-A1")
```


Nest by colony 
```{r}
#Nest by colony 
pi_nested_tp1 <- pi_tp1_norm %>%
  dplyr::select(SampleID, Light_Value, Slope_norm) %>%  # or whatever your P column is
  dplyr::rename(I = Light_Value, P = Slope_norm) %>%
  group_by(SampleID) %>%
  nest()
```

Calculate pi metrics 
```{r}
# Define a function to fit the model
fit_pi_model <- function(df) {
  tryCatch({
    fit <- nlsLM(P ~ Pmax * tanh(alpha * I / Pmax) + Rd,
                 data = df,
                 start = list(Pmax = max(df$P), alpha = 0.01, Rd = -0.1))
    pars <- coef(fit)
    tibble(Pmax = pars["Pmax"],
           alpha = pars["alpha"],
           Rd = pars["Rd"],
           Ik = pars["Pmax"]/pars["alpha"],
           Pgross = pars["Pmax"] + pars["Rd"])
  }, error = function(e) tibble(Pmax = NA, alpha = NA, Rd = NA, Ik = NA, Pgross = NA))
}

# Apply the model to each colony
pi_results <- pi_nested_tp1 %>%
  mutate(params = map(data, fit_pi_model)) %>%
  select(-data) %>%
  unnest(cols = params)

# Pivot pi data to long format for plotting 
pi_long <- pi_results %>%
  pivot_longer(cols = c(Pmax, alpha, Rd, Ik, Pgross),
               names_to = "Parameter",
               values_to = "Value")

# Merge with sample manifest info 
pi_long <- pi_long %>%
  left_join(meta, by = "SampleID")
```

Plot 
```{r}
ggplot(pi_long, aes(x = Phenotype, y = Value, fill = Phenotype)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.8) +
  facet_grid(Parameter ~ Species+Treatment, scales = "free_y") +
  scale_fill_manual(
    values = c("Resistant" = "#4b294fff", "Susceptible" = "#f3a17dff")) +
  theme_bw(base_size = 14) +
  labs(x = "Phenotype", y = "Value") +
  theme(legend.position = "none")
```



