---
title: "Rainbow tank temperature and light"
author: "Jill Ashey"
date: "2026-05-08"
output: html_document
---

This script plots temperature and light data collected from the rainbow tank monitoring system.

```{r}
library(tidyverse)
library(readr)
library(stringr)
library(ggplot2)
library(zoo) 
```

Read in metadata 
```{r}
tank_meta <- read.csv("../data/environmental/rainbow/Tank_metadata.csv")
```

## Temperature - rainbow tanks 

Read in temperature data 
```{r}
# Set path and tank numbers 
base_path <- "../data/environmental/rainbow/temperature/"
file_numbers <- 7:12

# Read and process files
all_temp <- file_numbers %>%
  map_df(function(num) {
    file_name <- paste0(base_path, num, "-data-2026-05-07.csv")
    temp_df <- read_csv(file_name)
    rainbow_col <- names(temp_df)[2]
    rainbow_num <- as.numeric(str_extract(rainbow_col, "\\d+"))
    temp_df %>%
      rename(DateTime = 1, temp_raw = 2) %>%
      mutate(
        Temperature_C = as.numeric(str_remove(temp_raw, " °C")),
        Rainbow_Tank = rainbow_num
      ) %>%
      select(DateTime, Temperature_C, Rainbow_Tank)
  })

# Make sure Rainbow_Tank is numeric 
tank_meta$Rainbow_Tank <- as.numeric(tank_meta$Rainbow_Tank)

all_temp <- all_temp %>%
  left_join(tank_meta, by = "Rainbow_Tank")
```

Filter by specific dates 
XXX

Plot by individual tank
```{r}
rainbow_tank_temp <- ggplot(all_temp, aes(x = DateTime, y = Temperature_C, color = Treatment_Tank)) +
  geom_line(alpha = 1) +
  labs(x = "Date and Time",
       y = "Temperature (°C)",
       color = "Tank ID") +
  theme_classic(); rainbow_tank_temp

ggsave(filename="../output/environmental/rainbow_tank_temperature.png", plot=rainbow_tank_temp, dpi=300, width=7, height=6, units="in")
```

Plot by treatment
```{r}
# Calculate mean + SE of treatment tanks 
se <- function(x) sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))

temp_plot_summary <- all_temp %>%
  group_by(DateTime, Treatment) %>%
  summarise(
    mean_temp = mean(Temperature_C, na.rm = TRUE),
    se_temp = se(Temperature_C),
    .groups = "drop"
  ) %>%
  group_by(Treatment) %>%
  mutate(
    temp_smooth = rollmean(mean_temp, k = 10, fill = NA)
  )

rainbow_trt_temp <- ggplot(temp_plot_summary, aes(x = DateTime, y = temp_smooth, color = Treatment, fill = Treatment)) +
  geom_ribbon(aes(ymin = temp_smooth - se_temp, 
                  ymax = temp_smooth + se_temp), 
              alpha = 0.3, color = NA) + 
  geom_line(size = 0.5) + # Thinner line
  scale_color_manual(values = c("Ambient" = "blue", "High" = "red")) +
  scale_fill_manual(values = c("Ambient" = "blue", "High" = "red")) +
  labs(x = "Date and Time",
       y = "Temperature (°C)",
       color = "Treatment") +
  theme_classic(); rainbow_trt_temp

ggsave(filename="../output/environmental/rainbow_treatment_temperature.png", plot=rainbow_trt_temp, dpi=300, width=7, height=6, units="in")
```

## Light - rainbow tanks 

Read in light data 
```{r}
# Set path and tank numbers 
base_path <- "../data/environmental/rainbow/light/"
file_numbers <- 7:12

# Read and process files
all_light <- file_numbers %>%
  map_df(function(num) {
    file_name <- paste0(base_path, num, "-data-2026-05-07.csv")
    light_df <- read_csv(file_name)
    controller_col <- names(light_df)[2]
    controller_num <- as.numeric(str_extract(controller_col, "\\d+"))
    light_df %>%
      rename(DateTime = 1, light_raw = 2) %>%
      mutate(
        Light_PAR = as.numeric(str_remove(light_raw, " PAR")),
        controller = controller_num
      ) %>%
      select(DateTime, Light_PAR, controller)
  })

# Make sure controller is numeric 
tank_meta$controller <- as.numeric(tank_meta$controller)

all_light <- all_light %>%
  left_join(tank_meta, by = "controller")
```

Correct for immersion effect (see details on [Apogee site](https://www.apogeeinstruments.com/underwater-par-measurements/)).
```{r}
all_light <- all_light %>%
  mutate(Light_PAR_corrected = Light_PAR * 1.25)
```

Filter by specific dates 
XXX

Plot by individual tank
```{r}
rainbow_tank_light <- ggplot(all_light, aes(x = DateTime, y = Light_PAR_corrected, color = Treatment_Tank)) +
  geom_line(alpha = 1) +
  labs(x = "Date and Time",
       y = "Light (PAR)",
       color = "Tank ID") +
  theme_classic(); rainbow_tank_light

ggsave(filename="../output/environmental/rainbow_tank_light.png", plot=rainbow_tank_light, dpi=300, width=7, height=6, units="in")
```

Plot by treatment
```{r}
# Calculate mean + SE of treatment tanks 
se <- function(x) sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))

light_plot_summary <- all_light %>%
  group_by(DateTime, Treatment) %>%
  summarise(
    mean_light = mean(Light_PAR_corrected, na.rm = TRUE),
    se_light = se(Light_PAR_corrected),
    .groups = "drop"
  ) %>%
  group_by(Treatment) %>%
  mutate(
    light_smooth = rollmean(mean_light, k = 10, fill = NA)
  )

rainbow_trt_light <- ggplot(light_plot_summary, aes(x = DateTime, y = light_smooth, color = Treatment, fill = Treatment)) +
  geom_ribbon(aes(ymin = light_smooth - se_light, 
                  ymax = light_smooth + se_light), 
              alpha = 0.3, color = NA) + 
  geom_line(size = 0.5) + # Thinner line
  scale_color_manual(values = c("Ambient" = "blue", "High" = "red")) +
  scale_fill_manual(values = c("Ambient" = "blue", "High" = "red")) +
  labs(x = "Date and Time",
       y = "Light (PAR)",
       color = "Treatment") +
  theme_classic(); rainbow_trt_light

ggsave(filename="../output/environmental/rainbow_treatment_light.png", plot=rainbow_trt_light, dpi=300, width=7, height=6, units="in")
```

## Analysis 

Test for differences between tanks 
```{r}
# Temperature
all_temp %>%
  aov(Temperature_C~Treatment_Tank, data=.)%>%
  summary() #%>%
  #TukeyHSD() %>%
  #tidy()

# Light
all_light %>%
  aov(Light_PAR_corrected~Treatment_Tank, data=.)%>%
  summary() #%>%
  #TukeyHSD() %>%
  #tidy()
```

Test for differences between treatments  
```{r}
# Temperature
all_temp %>%
  aov(Temperature_C~Treatment, data=.)%>%
  summary()

# Light
all_light %>%
  aov(Light_PAR_corrected~Treatment, data=.)%>%
  summary()
```
